\textbf{Supplementary Information}

### 2020-03-09 submission/manuscript.tex
Added OTU taxonomy.

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the ambient seawater were also processed for
comparison. The number of reads per sample ranged between 8,407 and
77,465 sequences (\autoref{supp-nseq_notus}). Even when the highest
sequencing effort was applied the rarefaction curves did not level off
that is a common observation in high-throughput 16S rRNA amplicon
sequencing approaches (\autoref{supp-rarefaction}). Following quality
curation and exclusion of sequences mentioned before reads were
clustered into 28,726 different OTUs at a similarity level of 97
\si{\percent}. Reads numbers were normalized to the minimum number of
sequences, 8,407 (\autoref{supp-nseq_notus}), through rarefaction
resulting in 17,007 different OTUs that ranged from 366 to 1,998 OTUs
per sample (\autoref{supp-calculators}). To determine seasonal changes
of richness and diversity the Observed Number of OTUs, Chao1, ACE,
Exponential Shannon (Jost, 2006) and Inverse Simpson were calculated
after normalization through rarefaction. Generally, richness estimators
and diversity indices showed similar trends. On average, higher values
were found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,688.5 ± 125.6 OTUs{]} and noninvaded {[}Number of OTUs, 1,744.8 ±
150.8 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,062.8 ± 209.6 OTUs) and lower values for picoplankton
communities in the ambient seawater (Number of OTUs, 528.1 ± 135.8 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{Cymodocea nodosa} communities showed a slow
increase towards the end of the study, while \emph{Caulerpa cylindracea}
(invaded and noninvaded) communities were characterized by slightly
larger values in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were found
at every time point. These OTUs made also a high proportion of total
sequences (39.1 -- 51.3 \si{\percent}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of
\emph{Cymodocea nodosa} associated communities indicated a separation
between Spring, Summer and Autumn/Winter samples (ANOSIM, R = 0.56,
\emph{p} \textless{} 0.01), while communities from the surfaces of
\emph{Caulerpa cylindracea} indicated a non so strongly supported, as in
previous cases, separation between Summer and Autumn/Winter/Spring
samples (ANOSIM, R = 0.31, \emph{p} \textless{} 0.05) (\autoref{pcoa}).

@@

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{Caulerpa cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{Cymodocea nodosa} communities were consisted of
only 0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{Caulerpa cylindracea} communities their proportion was
25.7 ± 11.3 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). The community was mainly consisted of no relative
\emph{Desulfobacteraceae}, \emph{Desulfatitalea}, no relative
\emph{Desulfobulbaceae}, \emph{Desulfobulbus}, no relative
\emph{Desulfocapsaceae}, \emph{Desulfopila}, \emph{Desulforhopalus},
\emph{Desulfotalea}, SEEP-SRB4 and uncultured \emph{Desulfocapsaceae}
(\autoref{desulfo}).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be find at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 51.3
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems, that
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{Cymodocea nodosa} and \emph{Caulerpa cylindracea}
associated communities were identified that could be linked to the
growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{Cymodocea nodosa} was characterized by a Spring
community during maximum seagrass proliferation, a Summer community
during a biomass maximum and a Autumn/Winter community during a biomass
decay. In contrast, \emph{Caulerpa cylindracea} started to proliferate
in late Spring and was characterized only by a Summer community during
maximal biomass increase and by a Autumn/Winter/Spring community when
the biomass were at the peak and the settlement started to subsequently
decay. Similar seasonal changes in the epiphytic community was described
also for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et
al.}, 2011). Higher temporal stability of \emph{Caulerpa cylindracea}
surface communities in comparison to \emph{Cymodocea nodosa} were also
observed in the higher proportion of shared communities between two
consecutive sampling points.

>>>>

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{Cymodocea nodosa} and macroalga \emph{Caulerpa
cylindracea} were analyzed. In addition, 18 samples (one of the samples
was sequenced two times) originating from picoplankton archaeal and
bacterial communities in the ambient seawater were also processed for
comparison. The number of reads per sample ranged between 8,410 and
77,465 sequences (\autoref{supp-nseq_notus}). Even when the highest
sequencing effort was applied the rarefaction curves did not level off
that is a common observation in high-throughput 16S rRNA amplicon
sequencing approaches (\autoref{supp-rarefaction}). Following quality
curation and exclusion of sequences mentioned before reads were
clustered into 28,729 different OTUs at a similarity level of 97
\si{\percent}. Reads numbers were normalized to the minimum number of
sequences, 8,410 (\autoref{supp-nseq_notus}), through rarefaction
resulting in 17,025 different OTUs that ranged from 351 to 2,059 OTUs
per sample (\autoref{supp-calculators}). To determine seasonal changes
of richness and diversity the Observed Number of OTUs, Chao1, ACE,
Exponential Shannon (Jost, 2006) and Inverse Simpson were calculated
after normalization through rarefaction. Generally, richness estimators
and diversity indices showed similar trends. On average, higher values
were found for \emph{Caulerpa cylindracea} (invaded {[}Number of OTUs,
1,694.5 ± 133.7 OTUs{]} and noninvaded {[}Number of OTUs, 1,738.2 ±
168.5 OTUs{]}), middle vales for \emph{Cymodocea nodosa} (Number of
OTUs, 1,057.7 ± 207.3 OTUs) and lower values for picoplankton
communities in the ambient seawater (Number of OTUs, 519 ± 139.9 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{Cymodocea nodosa} communities showed a slow
increase towards the end of the study, while \emph{Caulerpa cylindracea}
(invaded and noninvaded) communities were characterized by slightly
larger values in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{Caulerpa
cylindracea} (Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared
values were calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{Cymodocea nodosa}
and \emph{Caulerpa cylindracea} were approximately in the middle between
these two extremes. To assess seasonal changes in the proportion of
shared OTUs and communities the Jaccard's and Bray-Curtis Similarity
Coefficients were calculated between consecutive sampling points
(\autoref{shared}). Both coefficients showed similar trends. Temporal
proportional changes were more pronounced for seawater in comparison to
\emph{Cymodocea nodosa} and especially \emph{Caulerpa cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.1
\si{\percent} of OTUs from each surface associated community were found
at every time point. These OTUs made also a high proportion of total
sequences (35.4 -- 53.1 \si{\percent}). To further disentangle the
environmental and seasonal community dissimilarity a Principal
Coordinates Analysis (PCoA) based on Bray-Curtis distances and OTU
abundances was applied. It showed a clear separation between planktonic
and surface associated communities (\autoref{pcoa}). In addition, a
separation of epiphytic bacterial and archaeal communities based on host
species was determined. This separation was further supported by ANOSIM
(R = 0.96, \emph{p} \textless{} 0.001). Seasonal changes of
\emph{Cymodocea nodosa} associated communities indicated a separation
between Spring, Summer and Autumn/Winter samples (ANOSIM, R = 0.55,
\emph{p} \textless{} 0.01), while communities from the surfaces of
\emph{Caulerpa cylindracea} indicated a non so strongly supported, as in
previous cases, separation between Summer and Autumn/Winter/Spring
samples (ANOSIM, R = 0.32, \emph{p} \textless{} 0.01) (\autoref{pcoa}).

@@

In contrast to previously described high rank taxa,
\emph{Desulfobacterota} were specific to \emph{Caulerpa cylindracea}. On
average they were comprising 11.2 ± 13.3 \si{\percent} of all sequences.
While seawater and \emph{Cymodocea nodosa} communities were consisted of
only 0.1 ± 0.08 \si{\percent} and 1.0 ± 0.7 \si{\percent}
\emph{Desulfobacterota} sequences, respectively, in the invaded and
noninvaded \emph{Caulerpa cylindracea} communities their proportion was
25.7 ± 11.2 \si{\percent} and 24.0 ± 4.3 \si{\percent}, respectively
(\autoref{desulfo}). The community was mainly consisted of no relative
\emph{Desulfobacteraceae}, \emph{Desulfatitalea}, no relative
\emph{Desulfobulbaceae}, \emph{Desulfobulbus}, no relative
\emph{Desulfocapsaceae}, \emph{Desulfopila}, \emph{Desulforhopalus},
\emph{Desulfotalea}, SEEP-SRB4 and uncultured \emph{Desulfocapsaceae}
(\autoref{desulfo}).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be find at every
sampling time (\(\leq\) 1.1 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 53.1
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems, that
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{Cymodocea nodosa} and \emph{Caulerpa cylindracea}
associated communities were identified that could be linked to the
growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{Cymodocea nodosa} was characterized by a Spring
community during maximum seagrass proliferation, a Summer community
during a biomass maximum and a Autumn/Winter community during a biomass
decay. In contrast, \emph{Caulerpa cylindracea} started to proliferate
in late Spring and was characterized only by a Summer community during
maximal biomass increase and by a Autumn/Winter/Spring community when
the biomass were at the peak and the settlement started to subsequently
decay. Similar seasonal changes in the epiphytic community was described
also for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et
al.}, 2011). Higher temporal stability of \emph{Caulerpa cylindracea}
surface communities in comparison to \emph{Cymodocea nodosa} were also
observed in the higher proportion of shared communities between two
consecutive sampling points.

### 2020-03-09 submission/supplementary.tex
Added OTU taxonomy.

>>>>

\hypertarget{supplementary-table}{%
\subsection{Supplementary Table}\label{supplementary-table}}

\begingroup\fontsize{9}{11}\selectfont

\begin{longtable}[t]{>{\centering\arraybackslash}p{6em}ccccc}
\caption{\label{tab:nseq_notus}Sample ID, Community Type, Sampling Date and Season, No. of Sequences and No. of OTUs of each sample. No. of Sequences and OTUs was calculated after the exclusion of eukaryotic, chloroplast, mitochondrial and no relative sequences.\label{nseq_notus}}\\
\toprule
\textbf{Sample ID} & \textbf{Community Type} & \textbf{Date} & \textbf{Season} & \textbf{No. of Sequences} & \textbf{No. of OTUs}\\
\midrule
\endfirsthead
\caption[]{Sample ID, Community Type, Sampling Date and Season, No. of Sequences and No. of OTUs of each sample. No. of Sequences and OTUs was calculated after the exclusion of eukaryotic, chloroplast, mitochondrial and no relative sequences.\label{nseq_notus} \textit{(continued)}}\\
\toprule
\textbf{Sample ID} & \textbf{Community Type} & \textbf{Date} & \textbf{Season} & \textbf{No. of Sequences} & \textbf{No. of OTUs}\\
\midrule
\endhead
\
\endfoot
\bottomrule
\endlastfoot
3 & Seawater & 13 July 2017 & Summer & 26,008 & 824\\
5 & Seawater & 27 July 2017 & Summer & 58,955 & 993\\
7 & Seawater & 10 August 2017 & Summer & 32,626 & 852\\
9 & Seawater & 24 August 2017 & Summer & 60,929 & 1,058\\
11 & Seawater & 19 September 2017 & Summer & 46,111 & 1,171\\
13 & Seawater & 12 October 2017 & Autumn & 30,937 & 976\\
15 & Seawater & 23 November 2017 & Autumn & 27,581 & 1,183\\
17 & Seawater & 14 December 2017 & Autumn & 44,595 & 2,267\\
19 & Seawater & 13 February 2018 & Winter & 44,186 & 1,309\\
21 & Seawater & 27 March 2018 & Winter & 46,307 & 1,559\\
23a & Seawater & 24 April 2018 & Spring & 30,970 & 998\\
23b & Seawater & 24 April 2018 & Spring & 38,560 & 1,200\\
25 & Seawater & 22 May 2018 & Spring & 53,880 & 1,783\\
27 & Seawater & 19 June 2018 & Spring & 77,465 & 1,297\\
29 & Seawater & 10 July 2018 & Summer & 50,794 & 1,451\\
31 & Seawater & 9 August 2018 & Summer & 39,049 & 1,129\\
33 & Seawater & 4 September 2018 & Summer & 36,206 & 1,205\\
35 & Seawater & 5 October 2018 & Autumn & 49,580 & 1,013\\
37 & \textit{Cymodocea nodosa} (Invaded) & 23 November 2017 & Autumn & 31,235 & 1,548\\
41 & \textit{Cymodocea nodosa} (Invaded) & 4 December 2017 & Autumn & 24,237 & 1,176\\
45 & \textit{Cymodocea nodosa} (Invaded) & 14 December 2017 & Autumn & 32,683 & 1,051\\
49 & \textit{Cymodocea nodosa} (Invaded) & 13 February 2018 & Winter & 9,087 & 1,218\\
52 & \textit{Cymodocea nodosa} (Invaded) & 27 March 2018 & Winter & 17,003 & 1,218\\
55 & \textit{Cymodocea nodosa} (Invaded) & 24 April 2018 & Spring & 42,640 & 2,054\\
58 & \textit{Cymodocea nodosa} (Invaded) & 22 May 2018 & Spring & 21,337 & 1,292\\
61 & \textit{Cymodocea nodosa} (Invaded) & 19 June 2018 & Spring & 31,735 & 2,104\\
64 & \textit{Cymodocea nodosa} (Invaded) & 10 July 2018 & Summer & 35,733 & 1,792\\
67 & \textit{Cymodocea nodosa} (Invaded) & 9 August 2018 & Summer & 26,362 & 2,119\\
70 & \textit{Cymodocea nodosa} (Invaded) & 4 September 2018 & Summer & 23,278 & 1,715\\
73 & \textit{Cymodocea nodosa} (Invaded) & 5 October 2018 & Autumn & 29,909 & 2,205\\
38 & \textit{Caulerpa cylindracea} (Invaded) & 23 November 2017 & Autumn & 36,321 & 2,889\\
42 & \textit{Caulerpa cylindracea} (Invaded) & 4 December 2017 & Autumn & 28,370 & 3,249\\
46 & \textit{Caulerpa cylindracea} (Invaded) & 14 December 2017 & Autumn & 34,720 & 3,054\\
53 & \textit{Caulerpa cylindracea} (Invaded) & 27 March 2018 & Winter & 28,696 & 2,472\\
56 & \textit{Caulerpa cylindracea} (Invaded) & 24 April 2018 & Spring & 34,773 & 3,054\\
59 & \textit{Caulerpa cylindracea} (Invaded) & 22 May 2018 & Spring & 23,397 & 2,706\\
62 & \textit{Caulerpa cylindracea} (Invaded) & 19 June 2018 & Spring & 36,477 & 3,299\\
65 & \textit{Caulerpa cylindracea} (Invaded) & 10 July 2018 & Summer & 18,485 & 2,188\\
68 & \textit{Caulerpa cylindracea} (Invaded) & 9 August 2018 & Summer & 31,940 & 3,103\\
71 & \textit{Caulerpa cylindracea} (Invaded) & 4 September 2018 & Summer & 29,281 & 3,160\\
74 & \textit{Caulerpa cylindracea} (Invaded) & 5 October 2018 & Autumn & 11,705 & 1,705\\
39 & \textit{Caulerpa cylindracea} (Nonnvaded) & 23 November 2017 & Autumn & 22,064 & 2,438\\
43 & \textit{Caulerpa cylindracea} (Nonnvaded) & 4 December 2017 & Autumn & 13,661 & 1,887\\
47 & \textit{Caulerpa cylindracea} (Nonnvaded) & 14 December 2017 & Autumn & 8,410 & 1,451\\
51 & \textit{Caulerpa cylindracea} (Nonnvaded) & 13 February 2018 & Winter & 11,676 & 1,901\\
54 & \textit{Caulerpa cylindracea} (Nonnvaded) & 27 March 2018 & Winter & 39,472 & 3,136\\
57 & \textit{Caulerpa cylindracea} (Nonnvaded) & 24 April 2018 & Spring & 20,298 & 2,822\\
60 & \textit{Caulerpa cylindracea} (Nonnvaded) & 22 May 2018 & Spring & 33,047 & 3,297\\
63 & \textit{Caulerpa cylindracea} (Nonnvaded) & 19 June 2018 & Spring & 41,833 & 3,968\\
66 & \textit{Caulerpa cylindracea} (Nonnvaded) & 10 July 2018 & Summer & 27,036 & 2,674\\
69 & \textit{Caulerpa cylindracea} (Nonnvaded) & 9 August 2018 & Summer & 26,737 & 3,130\\
72 & \textit{Caulerpa cylindracea} (Nonnvaded) & 4 September 2018 & Summer & 31,869 & 3,240\\
75 & \textit{Caulerpa cylindracea} (Nonnvaded) & 5 October 2018 & Autumn & 33,068 & 3,086\\*
\end{longtable}
\endgroup{}
